Static contract checking for Haskell

### 2011-12-02 paper/paper.tex
Paper: Updated to reflect the multiple unrollings implementation.

$$\trans{M - f}, \dtrans{f=e[f/\rec{f}]}, \ctrans{\rec{f} \in c} \vdash \ctrans{f \in c}$$

Where $M - f $ means the content of the module $M$ without $f$'s
definition and $f$'s contract. However, this translation is not
complete: some proofs need to unroll the definition more than
once!\note[nc]{For a simple example, consider a contract that is
  proved by a finite number of reductions, but not by induction.  E.g.
}
  \begin{verbatim}
  isZero y = case y of Z -> True | S _ -> False
  eqNat x y = case x of
    Z -> isZero y
    S x' -> eqNatAux x' y
  eqNatAux x y = case y of
    Z -> False
    S y' -> eqNat x y'

@@

  OK, so we might hope to simply fully define $\rec{add}$, so that
  computations with $\rec{add}$ can be reduced.  That approach is
  unsound.  It's obvious, because as part of the translation of $f :::
  c$ we get $\rec{f} ::: c$ as an axiom, and so if $f ::: c$ doesn't
  actually hold, then $\rec{f} ::: c$ is contradictory.  For
  concreteness, here's simple example. Consider $f$ as above, but
  change the contract to $c := \cfc \to \{y: not ~y\}$.  Then $f$
  doesn't satisfy $c$, and a context that assumes $\rec{f}$ satisfies
  $c$ is contradictory when $\rec{f}$ is fully defined, by
  e.g. reducing $\rec{f}~Z = \true$.
 }
\subsection{Module checking}
A module is a collection of function definitions, data definitions and
contracts. What we want to do is to check that functions satisfy their
contract(s).

>>>>

$$\trans{M - f}, \dtrans{f=e[f/\rec{f}]}, \ctrans{\rec{f} \in c[\rec{f}/f]}^- \vdash \ctrans{f \in c}^+$$
\note[nc]{The LHS used to have $\ctrans{\rec{f} \in c}^-$, 
not $\ctrans{\rec{f} \in c[\rec{f}/f]}^-$, 
but the code has always implemented the latter. Which is preferable???}
Where $M - f $ means the content of the module $M$ without $f$'s
definition and $f$'s contract. However, this translation is not
complete: some proofs need to unroll the definition more than
once!\note[nc]{For a simple example, consider a contract that is
  proved by a finite number of reductions, but not by induction.  E.g.
}
  \begin{verbatim}
  isZero y = case y of Z -> True | S _ -> False
  eqNat x y = case x of
    Z -> isZero y
    S x' -> eqNatAux x' y
  eqNatAux x y = case y of
    Z -> False
    S y' -> eqNat x y'

@@

  OK, so we might hope to simply fully define $\rec{add}$, so that
  computations with $\rec{add}$ can be reduced.  That approach is
  unsound.  It's obvious, because as part of the translation of $f :::
  c$ we get $\rec{f} ::: c$ as an axiom, and so if $f ::: c$ doesn't
  actually hold, then $\rec{f} ::: c$ is contradictory.  For
  concreteness, here's simple example. Consider $f$ as above, but
  change the contract to $c := \cfc \to \{y: not ~y\}$.  Then $f$
  doesn't satisfy $c$, and a context that assumes $\rec{f}$ satisfies
  $c$ is contradictory when $\rec{f}$ is fully defined, by
  e.g. reducing $\rec{f}~Z = \true$.
 }

 So, we provide an option to control the number, $N$, of unrollings.
 First, note that recursive anonymous functions, and hence unrolling,
 only apply to checked functions in the SCC; the dependencies of the
 SCC being checked are not anonymized.  If $fs$ are the functions
 being checked, then for each $f xs = e \in fs$ we generate
 \[
   \bigwedge_{i=0,N+1} \dtrans{f_i ~xs = e[g_{i+1}/g]_{g \in fs}},
 \]
 where $g_0 = g$, $g_{N+1} = \rec{g}$, and $\{g_i\}_{i = 1 \text{ to } N }$ 
are new symbols, for each $g \in fs$.  Note that all functions
in $fs$ receive the same amount, $N$ levels, of unrolling.  We could
provide a per function unrolling value, but that would complicate
things, and it's not clear there would be any benefit.  The only
concern with the uniform unrolling is the complexity of the generated
theory.

\subsection{Module checking}
A module is a collection of function definitions, data definitions and
contracts. What we want to do is to check that functions satisfy their
contract(s).
